---
title: "Embedding polyviz charts in papers and documents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Embedding polyviz charts in papers and documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

polyviz charts live in the browser, but papers, theses, and slide decks
need files. This vignette covers the whole path from widget to page:
the four export formats of `pv_save()`, how to size them for print, the
knit path that makes Word and PDF documents work with no extra code, and
the download button on every rendered chart. None of the code below runs
while the vignette builds — most of it needs a local Chrome — so copy
the chunks into your own session.

The running example:

```{r example}
library(polyviz)

agg <- aggregate(revenue ~ region, pv_sales, sum)
w <- pv_bar(agg, "region", "revenue", title = "Revenue by region")
```

Printed at the console, `w` opens interactively in the Viewer. Note that
this only happens in an interactive session — `print(w)` inside a
non-interactive `Rscript` displays nothing, because there is no viewer
to open. In a script, save the chart instead.

## The four formats

`pv_save()` picks the format from the file extension alone. For `.png`,
`.svg`, and `.pdf` the chart is rendered for real: headless Chrome loads
the same d3 pipeline the interactive widget uses, the entrance animation
is switched off and the colour mode is pinned (light, unless you ask
otherwise), and the settled result is captured once nothing is still
drawing. A capture can never show a half-grown animation or inherit your
machine's dark theme. These three formats need the chromote package and
a Chrome-based browser; `.html` needs neither.

### PNG — Word, Google Docs, slides

```{r png}
pv_save(w, "revenue.png")
```

A raster image at `scale` times the pixel size — the default `scale = 2`
is a crisp "retina" capture. The right choice wherever the destination
re-compresses images or cannot place vector graphics.

### SVG — the web and vector editing

```{r svg}
pv_save(w, "revenue.svg")
```

One standalone SVG document with the title, legend, plot, and source
line all included as vector shapes and real text. It scales without
limit, and every element stays editable in Inkscape or Illustrator.

### PDF — LaTeX

```{r pdf}
pv_save(w, "revenue.pdf")
```

A true vector PDF from Chrome's print engine, sized exactly to the
chart with the font embedded — drop it straight into
`\includegraphics{revenue.pdf}`.

### HTML — sharing the interactive version

```{r html}
pv_save(w, "revenue.html")
```

The interactive widget as one self-contained file: every script,
stylesheet, and font inlined, tooltips and the entrance animation
intact. Email it, or put it on any static host. This format is built
purely in R — it is the only one that works without Chrome, and unlike
`htmlwidgets::saveWidget(selfcontained = TRUE)` it needs no pandoc
either.

## Sizing and scale for print

`width` and `height` are CSS pixels, and print maps them at 96 pixels
per inch. A chart meant to fill a 6.5-inch text block is therefore
624 pixels wide:

```{r sizing}
pv_save(w, "revenue.pdf", width = 624, height = 390)
```

`height = NULL` (the default) uses the fixed height the widget was built
with, if any, and 560 otherwise.

The chart lays itself out for the CSS size — margins, label density,
automatic orientation all react to it — so choose `width` and `height`
for the shape you want on the page, and let `scale` carry the
resolution. For a 300-dpi PNG, that is `scale = 300 / 96`:

```{r print-scale}
pv_save(w, "revenue.png", width = 624, height = 390, scale = 300 / 96)
```

Doubling `width` instead would sharpen nothing: it renders a genuinely
wider chart, whose type then looks half as big on the page. SVG and PDF
are resolution-independent and ignore `scale` altogether.

## Embedded fonts, or smaller files

By default (`embed_fonts = TRUE`) the bundled Inter font is inlined into
`.svg` and `.html` output as a base64 data URI, so the file renders with
the right type on any machine, in any viewer. The cost is size: the
font adds roughly half a megabyte. `embed_fonts = FALSE` keeps the file
small and falls back to a locally installed Inter (or the system font) —
a reasonable trade when you post-process the SVG on a machine that has
Inter installed anyway. `.png` and `.pdf` always carry their type and
ignore the argument.

## Knitting to Word and PDF

A chart dropped into an R Markdown document needs no special treatment.
When the document becomes HTML, the live interactive widget is embedded,
as always. For every static format — Word, PDF via LaTeX, plain or
GitHub-flavoured markdown, epub — the chart is rendered by `pv_save()`
to a PNG at the chunk's figure size and included like any other figure.

One line of setup belongs in the document, because knitr otherwise
reaches for its own generic widget screenshotter (the webshot package)
before asking the chart how it wants to be printed:

```{r knit-setup}
knitr::opts_chunk$set(screenshot.force = FALSE)
```

After that, an ordinary chunk does the right thing in every output
format. The chunk's `fig.width` and `fig.height` (inches) set the
chart's size, and `dpi` sets the capture resolution — knitr's default
dpi is screen resolution, so ask for more in a document headed for
print:

````
```{r revenue, fig.width = 6.5, fig.height = 4, dpi = 300}`r ''`
pv_bar(agg, "region", "revenue", title = "Revenue by region")
```
````

Charts built with `mode = "auto"` print light; a chart forced dark with
`mode = "dark"` stays dark. The static path has the same requirements
as `pv_save()`: the chromote package and a Chrome-based browser on the
machine doing the knitting.

## The download button

Every rendered chart carries a small arrow button in its top-right
corner, shown while the pointer (or keyboard focus) is on the chart and
permanently on touch screens. It offers the chart as a standalone SVG
or a 2x PNG, named after the chart's title — readers of a report or a
Shiny app can take any chart with them without asking you for the file.
The control never appears inside a saved or captured chart. Turn it off
per chart when it would distract:

```{r downloads}
pv_bar(agg, "region", "revenue") |>
  pv_downloads(FALSE)
```

## Troubleshooting

* **The saved image is dark (or light) when you meant the opposite.**
  A capture that went through `pv_save()` always pins the mode — light
  unless you passed `mode = "dark"` — so an unexpected theme means the
  explicit mode was lost on the way: usually knitr's own screenshotter
  grabbed the live widget, which follows the machine's theme, because
  `screenshot.force = FALSE` was missing from the setup chunk.
* **The chart is missing from the knitted Word or PDF document** (or an
  error naming chromote stands in its place). The knitting machine has
  no chromote package or no Chrome-based browser; install both. Only
  the `.html` format works without them.
* **pandoc complaints.** pandoc only matters for
  `htmlwidgets::saveWidget(selfcontained = TRUE)`; `pv_save()` builds
  its self-contained `.html` without pandoc, so use it instead.
* **A capture looks half-loaded.** `pv_save()` waits for the chart to
  settle, but a chart that keeps fetching things can outlast the wait —
  raise `delay`. If a JavaScript error broke the render, `pv_save()`
  warns quoting the error; treat that warning as the explanation.
