---
title: "Embedding polyviz charts in papers and documents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Embedding polyviz charts in papers and documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

polyviz charts live in the browser, but papers, theses, and slide decks
need files. This vignette covers the whole path from widget to page:
the five export formats of `pv_save()`, how to size them for print, the
paper theme and texture fills that keep a figure legible in greyscale,
Swiss locale formatting, the knit path that makes Word and PDF documents
work with no extra code, the alt text those documents carry, whole
PowerPoint decks with `pv_deck()`, and the download button on every
rendered chart. None of the code below runs while the vignette builds —
most of it needs a local Chrome — so copy the chunks into your own
session.

The running example:

```{r example, eval = FALSE}
library(polyviz)

agg <- aggregate(revenue ~ region, pv_sales, sum)
w <- pv_bar(agg, "region", "revenue", title = "Revenue by region")
```

Printed at the console, `w` opens interactively in the Viewer. Note that
this only happens in an interactive session — `print(w)` inside a
non-interactive `Rscript` displays nothing, because there is no viewer
to open. In a script, save the chart instead.

## The five formats

`pv_save()` picks the format from the file extension alone. For `.png`,
`.svg`, and `.pdf` — and every frame of a `.gif` — the chart is
rendered for real: headless Chrome loads
the same d3 pipeline the interactive widget uses, the entrance animation
is switched off and the colour mode is pinned (light, unless you ask
otherwise), and the settled result is captured once nothing is still
drawing. A capture can never show a half-grown animation or inherit your
machine's dark theme. The capture formats need the chromote package and
a Chrome-based browser (`.gif` additionally needs gifski); `.html`
needs neither.

### PNG — Word, Google Docs, slides

```{r png, eval = FALSE}
pv_save(w, "revenue.png")
```

A raster image at `scale` times the pixel size — the default `scale = 2`
is a crisp "retina" capture. The right choice wherever the destination
re-compresses images or cannot place vector graphics.

### SVG — the web and vector editing

```{r svg, eval = FALSE}
pv_save(w, "revenue.svg")
```

One standalone SVG document with the title, legend, plot, and source
line all included as vector shapes and real text. It scales without
limit, and every element stays editable in Inkscape or Illustrator.

### PDF — LaTeX

```{r pdf, eval = FALSE}
pv_save(w, "revenue.pdf")
```

A true vector PDF from Chrome's print engine, sized exactly to the
chart with the font embedded — drop it straight into
`\includegraphics{revenue.pdf}`.

### HTML — sharing the interactive version

```{r html, eval = FALSE}
pv_save(w, "revenue.html")
```

The interactive widget as one self-contained file: every script,
stylesheet, and font inlined, tooltips and the entrance animation
intact. Email it, or put it on any static host. This format is built
purely in R — it is the only one that works without Chrome, and unlike
`htmlwidgets::saveWidget(selfcontained = TRUE)` it needs no pandoc
either.

### GIF — the bar-chart race, animated

```{r gif, eval = FALSE}
r <- pv_race(pv_city_population, time = "year", id = "city",
             value = "population", top_n = 8)
pv_save(r, "race.gif")
```

A looping animated GIF of the race, for slide decks, chat, and READMEs
(needs the gifski package). Frames are rendered one by one at exact
positions on the race's timeline — never screenshotted off the running
animation — so the file comes out identical on every run. The chart's
own `duration` sets the tempo, `fps` the frame rate (default 20), and
the starting order and final standings each hold briefly so the loop
reads. Only `pv_race()` charts save as `.gif`: every other chart merely
animates its entrance, and a `.png` shows the same finished chart —
asking for a GIF of one is an error saying so.

## Sizing and scale for print

`width` and `height` are CSS pixels, and print maps them at 96 pixels
per inch. A chart meant to fill a 6.5-inch text block is therefore
624 pixels wide:

```{r sizing, eval = FALSE}
pv_save(w, "revenue.pdf", width = 624, height = 390)
```

`height = NULL` (the default) uses the fixed height the widget was built
with, if any, and 560 otherwise.

The chart lays itself out for the CSS size — margins, label density,
automatic orientation all react to it — so choose `width` and `height`
for the shape you want on the page, and let `scale` carry the
resolution. For a 300-dpi PNG, that is `scale = 300 / 96`:

```{r print-scale, eval = FALSE}
pv_save(w, "revenue.png", width = 624, height = 390, scale = 300 / 96)
```

Doubling `width` instead would sharpen nothing: it renders a genuinely
wider chart, whose type then looks half as big on the page. SVG and PDF
are resolution-independent and ignore `scale` altogether.

## Embedded fonts, or smaller files

By default (`embed_fonts = TRUE`) the bundled Inter font is inlined into
`.svg` and `.html` output as a base64 data URI, so the file renders with
the right type on any machine, in any viewer. The cost is size: the
font adds roughly half a megabyte. `embed_fonts = FALSE` keeps the file
small and falls back to a locally installed Inter (or the system font) —
a reasonable trade when you post-process the SVG on a machine that has
Inter installed anyway. `.png` and `.pdf` always carry their type and
ignore the argument.

## Greyscale safety: the paper theme and textures

Many journals print figures in greyscale, and photocopiers always do.
Two pieces pair up to keep a chart legible with every hue thrown away:

```{r paper-theme, eval = FALSE}
pv_set_theme(pv_theme_paper(serif = TRUE))

mix <- aggregate(revenue ~ region + product, pv_sales, sum)
pv_bar(mix, x = "region", y = "revenue", series = "product",
       stack = "stack", title = "Revenue by region and product") |>
  pv_textures() |>
  pv_save("figure-3.pdf", width = 624, height = 390)

pv_reset_theme()
```

`pv_theme_paper()` is a complete print-first theme: a pure white
surface, neutral near-black inks, sequential and diverging ramps
re-anchored to match, and a five-colour categorical palette whose slots
climb a luminance ladder — adjacent series stay tellable apart as pure
greys. `serif = TRUE` sets the widget type in a serif stack for
journals that want figures to match their text. Like any theme it is
session-wide: set it, build the figures, then `pv_reset_theme()`.

`pv_textures()` adds the belt to those braces on any chart with filled
marks (bar, area, donut, treemap): each series is hatched in its own
diagonal and line spacing over a lightened solid of its colour, so
series identity is carried by pattern as well — an encoding that
survives greyscale print and colour-vision deficiency alike. Legend
swatches wear the same texture, and every export format carries it
along.

## Swiss number and date formats

A figure in a German-language paper should not write `10,000` where the
reader expects `10'000`. `pv_locale()` switches every chart built
afterwards to one of the packaged Swiss conventions — `"de-CH"`,
`"fr-CH"`, `"it-CH"`, or `"en-CH"`:

```{r locale, eval = FALSE}
pv_locale("de-CH")

pop <- subset(pv_city_population, year == 2024)
pop <- head(pop[order(-pop$population), ], 8)
pv_bar(pop, x = "city", y = "population", sort = TRUE,
       title = "Wo die städtische Schweiz wohnt") |>
  pv_save("bevoelkerung.pdf")

pv_locale(NULL)
```

On the saved figure the population axis reads `100'000` through
`400'000` — apostrophe-grouped in full, the Federal Chancellery's
convention — and tooltips in the interactive version follow suit;
`"fr-CH"` groups with a space and writes the decimal comma instead, and
each locale carries its language's month and weekday names for date
axes. Ticks below 10'000 stay ungrouped, so year axes keep reading
`2024`. The setting travels inside each widget's payload, so a chart
keeps its locale wherever it goes, and `pv_locale(NULL)` puts d3's
stock US-English formatting back — with no locale set, charts render
exactly as they always have.

## Knitting to Word and PDF

A chart dropped into an R Markdown document needs no special treatment.
When the document becomes HTML, the live interactive widget is embedded,
as always. For every static format — Word, PDF via LaTeX, plain or
GitHub-flavoured markdown, epub — the chart is rendered by `pv_save()`
to a PNG at the chunk's figure size and included like any other figure.

One line of setup belongs in the document, because knitr otherwise
reaches for its own generic widget screenshotter (the webshot package)
before asking the chart how it wants to be printed:

```{r knit-setup, eval = FALSE}
knitr::opts_chunk$set(screenshot.force = FALSE)
```

After that, an ordinary chunk does the right thing in every output
format. The chunk's `fig.width` and `fig.height` (inches) set the
chart's size, and `dpi` sets the capture resolution — knitr's default
dpi is screen resolution, so ask for more in a document headed for
print:

````
```{r revenue, fig.width = 6.5, fig.height = 4, dpi = 300}`r ''`
pv_bar(agg, "region", "revenue", title = "Revenue by region")
```
````

Charts built with `mode = "auto"` print light; a chart forced dark with
`mode = "dark"` stays dark. The static path has the same requirements
as `pv_save()`: the chromote package and a Chrome-based browser on the
machine doing the knitting.

## Alt text in knitted documents

Every chart writes a short description of itself when it is built — the
chart form in plain words, the title, what is measured, and the data's
actual shape. In the browser that text reaches screen readers through
the widget; in a knitted document it rides along as the included
figure's `fig.alt`, so a knit to markdown, GitHub-flavoured markdown,
or epub keeps an accessible description on the image with no extra
code. (Word and LaTeX keep their usual figure markup — knitr declares
`fig.alt` unsupported there — and an author's own `fig.alt` or
`fig.cap` chunk option always wins.)

`pv_alt_text()` returns the generated text, which also makes a solid
first draft of a figure caption; `pv_alt()` replaces it with your own
words wherever the data deserves a better sentence:

```{r alt-text, eval = FALSE}
pv_alt_text(w)  # what the chart will say about itself

w |> pv_alt("Revenue by region: North leads, the other three are close.")
```

## Charts to slides

`pv_deck()` skips the copy-paste round through PowerPoint entirely:
give it a list of charts and it writes a `.pptx`, one chart per slide.

```{r deck, eval = FALSE}
deck <- list(
  pv_bar(agg, "region", "revenue", title = "Revenue by region"),
  pv_donut(agg, "region", "revenue", title = "Share of revenue")
)
pv_deck(deck, "review.pptx", title = "Quarterly review")
```

Every slide is a real render through the same pipeline as `pv_save()`,
captured at exactly the slide's dimensions and placed edge to edge on a
blank layout — what lands in the deck is pixel for pixel what a
`pv_save()` capture shows, at 2x resolution by default. Slides are 16:9
(`ratio = "4:3"` switches, or give `width`/`height` in inches for
anything else), `title` opens the deck with a title slide on the same
chart surface, and each chart's title and source line go into its
slide's speaker notes so the deck stays searchable. Needs the officer
package to write the file, plus chromote and Chrome to render — and it
composes with everything above: build the deck under
`pv_theme_paper()`, a locale, or with textures, and the slides carry
the look.

## The download button

Every rendered chart carries a small arrow button in its top-right
corner, shown while the pointer (or keyboard focus) is on the chart and
permanently on touch screens. It offers the chart as a standalone SVG
or a 2x PNG, named after the chart's title — readers of a report or a
Shiny app can take any chart with them without asking you for the file.
The control never appears inside a saved or captured chart. Turn it off
per chart when it would distract:

```{r downloads, eval = FALSE}
pv_bar(agg, "region", "revenue") |>
  pv_downloads(FALSE)
```

## Troubleshooting

* **The saved image is dark (or light) when you meant the opposite.**
  A capture that went through `pv_save()` always pins the mode — light
  unless you passed `mode = "dark"` — so an unexpected theme means the
  explicit mode was lost on the way: usually knitr's own screenshotter
  grabbed the live widget, which follows the machine's theme, because
  `screenshot.force = FALSE` was missing from the setup chunk.
* **The chart is missing from the knitted Word or PDF document** (or an
  error naming chromote stands in its place). The knitting machine has
  no chromote package or no Chrome-based browser; install both. Only
  the `.html` format works without them.
* **pandoc complaints.** pandoc only matters for
  `htmlwidgets::saveWidget(selfcontained = TRUE)`; `pv_save()` builds
  its self-contained `.html` without pandoc, so use it instead.
* **A capture looks half-loaded.** `pv_save()` waits for the chart to
  settle, but a chart that keeps fetching things can outlast the wait —
  raise `delay`. If a JavaScript error broke the render, `pv_save()`
  warns quoting the error; treat that warning as the explanation.
