---
title: "Get started with polyviz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with polyviz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, eval = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

polyviz draws interactive D3 charts from plain R data frames — no
JavaScript, no grammar to learn, one function call per chart. This is
the five-minute tour: a first chart, a second, the automatic behaviours
and how to override them, layers, theming, and the way out to a printed
page. The charts open in your Viewer as you go, so paste along; none of
the chunks run while the vignette builds.

## Install

```{r install, eval = FALSE}
# install.packages("devtools")
devtools::install_github("jastephan63/polyviz")
```

Everything the charts need ships in the package, bundled D3 included.
Python is optional (profiling falls back to pure R without it), and
headless Chrome is only needed at the very end, for `pv_save()`.

## A first chart, in one line

Fourteen real Swiss open-government datasets are bundled. `pv_weather`
holds six years of daily weather in Lucerne, from MeteoSwiss — hand it
straight to a line chart:

```{r first-chart, eval = FALSE}
library(polyviz)

pv_line(pv_weather, x = "date", y = "temp_mean")
```

The chart draws itself in, follows your light/dark setting, re-renders
when its container resizes, and a crosshair tooltip reads out the exact
value under the cursor. That is the whole interface: a data frame,
column names, options.

## Add a title and a source line

Every chart shares the same heading options — `title`, `subtitle`, and
a small grey `source` credit at the bottom left, newsroom style:

```{r titled, eval = FALSE}
pv_line(pv_weather, x = "date", y = "temp_mean",
        title = "Six years of Lucerne weather",
        subtitle = "Daily mean temperature",
        source = "Source: MeteoSwiss")
```

## A second chart type

All 34 chart types work the same way. Bars want one row per bar, so
aggregate first — and invalid data fails with a clear R-side message
rather than a broken chart:

```{r second-chart, eval = FALSE}
sales <- aggregate(revenue ~ region, pv_sales, sum)
pv_bar(sales, x = "region", y = "revenue", title = "Revenue by region")
```

## The "auto" defaults, and overriding them

Charts adapt to their data and their pixels at render time. Here
fifteen municipality names are too long to sit side by side under
vertical bars, so `horizontal = "auto"` — the default — flips the chart
sideways: labels upright down the left, the exact value at each bar's
end:

```{r auto, eval = FALSE}
f25 <- subset(pv_fiscal, year == 2025)
top15 <- head(f25[order(-f25$equalization_chf), ], 15)

pv_bar(top15, x = "municipality", y = "equalization_chf",
       title = "Largest equalization payments, 2025",
       source = "Source: LUSTAT Statistik Luzern")
```

Every automatic behaviour has an explicit `TRUE`/`FALSE`/`"auto"`
override. Force the vertical layout back and you can see what the
default saved you from — fifteen labels fighting for the same row:

```{r override, eval = FALSE}
pv_bar(top15, x = "municipality", y = "equalization_chf",
       horizontal = FALSE)
```

## Layer with the pipe

Modifiers pipe onto an existing chart. `pv_trend()` fits a smoother
with a confidence ribbon:

```{r trend, eval = FALSE}
pv_scatter(f25, x = "resource_index", y = "equalization_chf") |>
  pv_trend("loess")
```

The same shape gives you small multiples (`pv_facet()`), annotations
(`pv_annotate()`), print-safe hatching (`pv_textures()`), and linked
selection across charts (`pv_link()`).

## Theme and locale, in two lines

```{r theme-locale, eval = FALSE}
pv_set_theme(pv_theme_paper())  # print-first: white surface, luminance-ordered colours
pv_locale("de-CH")              # Swiss numbers (10'000) and German month names
```

Every chart built from here on wears both. `pv_set_theme()` also takes
your own brand colours — with an accessibility check on the way in —
and `pv_reset_theme()` and `pv_locale(NULL)` restore the defaults.

## Into a paper

`pv_save()` picks the format from the file extension — a crisp 2x PNG
for Word and slides, true vector PDF for LaTeX, standalone SVG, or one
self-contained interactive HTML file:

```{r save, eval = FALSE}
w <- pv_bar(sales, x = "region", y = "revenue",
            title = "Revenue by region")
pv_save(w, "revenue.png")
pv_save(w, "revenue.pdf")
```

The capture formats render the chart for real in headless Chrome (the
chromote package plus a Chrome-based browser). Knitting to Word or PDF
needs no call at all — a chart in a chunk becomes a print-quality
figure by itself. Sizing, alt text, animated GIFs, and whole PowerPoint
decks: `vignette("embedding-charts")`.

## Where to go next

```{r next, eval = FALSE}
pv_suggest(pv_weather)  # which charts fit this data frame?
pv_demo()               # every chart family, live, with controls
```

* The **[live gallery](https://jastephan63.github.io/polyviz/)** — all
  34 chart types, interactive, each explained and running on real Swiss
  open data, with the R code that made it.
* `pv_demo()` opens the same idea as a Shiny app on your machine, with
  live controls for each chart's main options.
* `pv_suggest()` inspects a data frame and prints runnable chart calls
  with the real column names filled in, each with a line of reasoning —
  every suggestion is proven to run before it is offered.
* The **[cheatsheet (PDF)](https://jastephan63.github.io/polyviz/polyviz-cheatsheet.pdf)**
  fits the whole package on a desk-side sheet.
* The **[reference site](https://jastephan63.github.io/polyviz/reference/)**
  documents every function; the other vignettes go deeper — exploring a
  dataset (`vignette("polyviz-eda")`), charts in papers
  (`vignette("embedding-charts")`), live Swiss open data
  (`vignette("swiss-open-data")`), and the SQL/SAS/Python backends
  (`vignette("polyglot-backend")`).
