---
title: "The polyglot backend"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The polyglot backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(polyviz)
```

You write R. Underneath, polyviz talks to whatever engine fits the job:
SQLite or DuckDB for data that lives in a database (`pv_db_connect()`
picks by file extension, or take `driver = "duckdb"` explicitly — a
DuckDB connection also queries Parquet and CSV files in place, larger
than memory), `haven` for SAS files and their variable labels, an
optional Python module for numeric profiling, and D3.js for every
chart. This vignette shows each seam — and makes the point that none
of them require you to leave R, or even to notice which engine did
the work.

## One reader for every file

`pv_read()` dispatches on the file extension: CSV and RDS via base R,
`.sas7bdat`/`.xpt` via haven, `.sqlite`/`.db` via DBI, and — with the
duckdb package installed — `.duckdb`/`.ddb` databases and `.parquet`
files via DuckDB. The package bundles a small SQLite database to
demonstrate:

```{r read-dispatch}
db <- system.file("extdata", "demo.sqlite", package = "polyviz")
sales <- pv_read(db, table = "sales")
head(sales, 3)
```

## SQL, when the data lives in a database

For anything beyond "give me the table", open a real connection and
push work to the database. Aggregating in SQL and charting the small
result in R is the natural division of labour:

```{r sql}
con <- pv_db_connect(db)
pv_db_tables(con)

by_region <- pv_query(con, "
  SELECT region, SUM(revenue) AS revenue
  FROM sales
  GROUP BY region
  ORDER BY revenue DESC
")
by_region
```

Parameters are bound with `?` placeholders — values never get pasted
into the SQL string:

```{r sql-params}
pv_query(con,
  "SELECT COUNT(*) AS n FROM sales WHERE region = ? AND units > ?",
  params = list("North", 300))
```

The result of a query is an ordinary data frame, so it flows straight
into a chart:

```{r sql-chart}
pv_bar(by_region, x = "region", y = "revenue",
       title = "Aggregated by SQLite, drawn by D3",
       subtitle = "Revenue by region, summed in the database")
```

Schema and seed scripts can stay as plain `.sql` files and run with
`pv_run_sql_file()`. Scratch databases work the same way —
`pv_db_connect()` with no path opens an in-memory one, and
`pv_db_write()` puts any data frame into it.

```{r sql-cleanup}
pv_db_disconnect(con)
```

## SAS, without SAS

`haven` reads and writes SAS files directly, so polyviz can round-trip
them — including *variable labels*, the SAS metadata that usually gets
lost in translation. Attach labels, write an `.xpt` transport file,
read it back:

```{r sas-roundtrip}
labelled <- pv_set_labels(mtcars[1:4], c(
  mpg = "Miles per US gallon",
  cyl = "Number of cylinders",
  disp = "Displacement (cu in)",
  hp = "Gross horsepower"
))

xpt <- tempfile(fileext = ".xpt")
pv_write_sas(labelled, xpt)

back <- pv_read_sas(xpt)
pv_labels(back)
```

The labels survive the trip, and downstream tools use them:
`pv_summary()` shows a label column whenever one is present.

```{r sas-summary}
pv_summary(back)
```

## Python, when it happens to be there

Numeric profiling (`pv_profile()`, `pv_outliers()`) can run on a small
Python module bundled with the package, via `reticulate`. The rule:
Python is never required. Every Python-backed function has a pure-R
implementation that returns identical results, and `engine = "auto"`
(the default) quietly picks whichever is available.

```{r python}
pv_py_available()

prof <- pv_profile(mtcars[c("mpg", "hp", "wt")])
attr(prof, "engine")
prof
```

On a machine with Python, that `engine` attribute reads `"python"`; on
one without, `"r"` — and the numbers are the same either way, which the
package's test suite asserts. Forcing a specific engine is explicit:
`engine = "r"` always works, `engine = "python"` errors (rather than
silently downgrading) when Python is missing.

```{r python-match}
identical(prof$mean, pv_profile(mtcars[c("mpg", "hp", "wt")],
                                engine = "r")$mean)
```

## JavaScript: what actually travels to D3

Every chart function builds a plain payload — data as row-wise records,
options, and the colour theme — and hands it to the bundled D3 v7
renderers through htmlwidgets. No JavaScript is written at chart time,
and you can inspect exactly what crosses the boundary:

```{r payload}
w <- pv_bar(by_region, x = "region", y = "revenue")
names(w$x)
str(w$x$data)
```

`type` picks the renderer, `theme` carries the full light *and* dark
token sets (the browser picks at render time, following the viewer's
system preference), and modifiers like `pv_trend()` just add fields to
this list: the loess or lm fit runs in R, and only the fitted points
travel to JavaScript. R stays the single source of statistical truth;
D3 only draws.

## Coming back: events and linked selections

The boundary runs both ways. Inside a Shiny app, every chart reports
interactions as input values named `<outputId>_<event>`:

```{r shiny-sketch, eval = FALSE}
library(shiny)

ui <- fluidPage(
  pvchartOutput("revenue"),
  verbatimTextOutput("detail")
)

server <- function(input, output, session) {
  output$revenue <- renderPvchart(
    pv_bar(by_region, x = "region", y = "revenue")
  )
  # Clicking a bar arrives here as input$revenue_click.
  output$detail <- renderPrint(input$revenue_click)
}

shinyApp(ui, server)
```

And without any server at all, `pv_link()` joins charts through a
crosstalk selection group — brushing points in one chart dims the
non-matching marks in every linked widget, even in a static HTML page:

```{r crosstalk-sketch, eval = FALSE}
f25 <- subset(pv_fiscal, year == 2025)
sd <- crosstalk::SharedData$new(f25)

# Both charts are built from the same rows and linked to the same
# group: selecting municipalities in the scatter highlights them on
# the map, and vice versa.
htmltools::browsable(htmltools::tagList(
  pv_scatter(f25, x = "resource_index", y = "equalization_chf",
             label = "municipality") |> pv_link(sd),
  pv_choropleth(f25, id = "municipality_id",
                value = "resource_index") |> pv_link(sd)
))
```

Four engines, one interface — and the seams only show when you go
looking for them.
