---
title: "Exploring a dataset with polyviz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring a dataset with polyviz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(polyviz)
```

polyviz is built around a simple loop: look at the data, ask a question,
draw the chart that answers it, repeat. This vignette walks that loop
once, end to end, on `pv_fiscal` — a real dataset that ships with the
package.

## The data: who pays for whom in Canton Lucerne

Like most Swiss cantons, Lucerne runs a *fiscal equalization* scheme:
municipalities with a strong tax base pay in, municipalities with a weak
one receive support. `pv_fiscal` holds the key figures for every Lucerne
municipality since 2020, published by LUSTAT Statistik Luzern:

* `resource_per_capita` — tax resource potential per resident, in CHF;
* `resource_index` — the same figure indexed so the cantonal average
  is 100;
* `equalization_chf` — the equalization amount the municipality
  receives (0 for net contributors).

First look, before any chart:

```{r summary}
pv_summary(pv_fiscal)
```

About 80 municipalities per year (municipalities occasionally merge, so
the count moves), no missing values, and years running from 2020 to
`r max(pv_fiscal$year)` — the last two of which are published
projections, not accounts. `pv_profile()` gives the numeric columns a
closer read:

```{r profile}
pv_profile(pv_fiscal[c("resource_per_capita", "resource_index",
                       "equalization_chf")], engine = "r")
```

Two things stand out. The resource index spans roughly 37 to 292 — the
strongest municipality has almost eight times the per-resident tax
resources of the weakest. And `equalization_chf` has a 25th percentile
of 0: a sizeable minority of municipalities receive nothing at all.

## How unequal is the tax base?

```{r histogram}
f25 <- subset(pv_fiscal, year == 2025)

pv_histogram(
  f25, x = "resource_index",
  title = "Most Lucerne municipalities sit well below the average",
  subtitle = "Resource index, 2025 (cantonal average = 100)",
  source = "Quelle: LUSTAT Statistik Luzern"
) |>
  pv_annotate(pv_vline(100, label = "cantonal average"))
```

The distribution is heavily right-skewed. The bulk of municipalities
cluster between roughly 50 and 90, below the average line; a long thin
tail of lakeside communities — Meggen, Weggis, Vitznau — stretches past
200. Because the index weights by population, a handful of strong
municipalities is enough to pull the average of 100 far above the
median (about `r round(median(f25$resource_index))` in 2025).

Is that a snapshot quirk or a trend? Boxplots by year show the whole
distribution moving:

```{r boxplot}
pv_boxplot(
  pv_fiscal, value = "resource_index", group = "year",
  title = "The middle is drifting away from the top",
  subtitle = "Resource index by year; 2026-27 are projections",
  source = "Quelle: LUSTAT Statistik Luzern"
)
```

The median slides from about 78 in 2020 to about 70 in the 2027
projection, while the maximum barely moves. On this measure, dispersion
is widening — which is exactly the situation equalization exists for.

## Who receives, and how much?

```{r scatter}
pv_scatter(
  f25, x = "resource_index", y = "equalization_chf",
  label = "municipality",
  title = "Payments fade out as the tax base approaches the average",
  subtitle = "Resource index vs. equalization received, 2025",
  ylab = "equalization (CHF)",
  source = "Quelle: LUSTAT Statistik Luzern"
) |>
  pv_trend("loess") |>
  pv_annotate(
    pv_vline(100, label = "cantonal average"),
    pv_note(65.9, 23278722, "Emmen", dx = -18, dy = 10)
  )
```

The broad shape is what the scheme promises: payments are concentrated
below the average and reach zero near it. In 2025,
`r sum(f25$equalization_chf == 0)` of `r nrow(f25)` municipalities
receive nothing, and none of them has an index under about 87.

```{r receiver-correlation, include = FALSE}
receivers <- subset(f25, equalization_chf > 0)
recv_cor <- round(cor(receivers$resource_index,
                      receivers$equalization_chf), 2)
```

But the scatter is honest about what the index alone does not explain.
The largest single payment — about CHF 23 million to Emmen — goes to a
municipality that is far from the poorest; Emmen is simply large, and
`equalization_chf` is a total, not a per-resident amount. Among
receivers, the correlation between index and amount is close to zero
(`r recv_cor`), because population size dominates. A ranking of
absolute amounts is therefore mostly a ranking of size:

```{r lollipop}
top20 <- head(f25[order(-f25$equalization_chf), ], 20)

pv_lollipop(
  top20, x = "municipality", y = "equalization_chf",
  title = "The biggest cheques go to the biggest municipalities",
  subtitle = "Largest equalization payments, 2025 (CHF)",
  source = "Quelle: LUSTAT Statistik Luzern"
)
```

## Where the weak tax bases are

The municipality numbers in `pv_fiscal` are official BFS ids, and the
bundled `pv_lucerne_map` boundaries carry the same ids — so the join to
a map is automatic:

```{r map}
pv_choropleth(
  f25, id = "municipality_id", value = "resource_index",
  title = "Tax resources follow the lake",
  subtitle = "Resource index by municipality, 2025 (average = 100)",
  source = "Quelle: LUSTAT Statistik Luzern; Grenzen: BFS ThemaKart"
)
```

The geography is stark: the strong tax bases hug Lake Lucerne and the
agglomeration, while the rural north-west — the Napf hinterland around
Romoos and Luthern — holds the weakest. Fiscal equalization in Lucerne
is, to a first approximation, a transfer from the lakeshore to the
hills.

## Where to go next

* The whole pipeline — CSV, SQLite, SAS, and the Python profiler —
  is covered in `vignette("polyglot-backend")`.
* `pv_report()` writes a one-call HTML report (summary, histograms,
  correlations) of any data frame, built from the same widgets used
  here.
* Every chart above is interactive in the browser: hover for exact
  values, and in Shiny each chart also reports clicks (see
  `?pvchartOutput`).
