---
title: "Swiss open data from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Swiss open data from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The datasets bundled with polyviz are snapshots — frozen on purpose, so
examples and tests give the same answer every time. But all of them come
from live, openly licensed Swiss government portals, and the package can
talk to those portals directly: search the federal catalogue, download a
dataset, and put it on a map, without leaving R. This vignette walks
those doors one by one and ends with the full path from portal to paper.

None of the chunks below run while the vignette builds — they all need
the network — so copy them into your own session.

```{r library, eval = FALSE}
library(polyviz)
```

## Three doors to the data

* **opendata.swiss** is the federal catalogue over (nearly) everything
  Swiss public bodies publish. `pv_search_opendata()` queries it,
  `pv_fetch_opendata()` downloads a dataset it lists.
* **stats.swiss** is the Federal Statistical Office's own dissemination
  platform. `pv_fetch_bfs()` pulls a dataflow from its SDMX API — the
  richest door when you know the BFS publishes what you want.
* **data.lustat.ch** hosts the open CSVs of LUSTAT Statistik Luzern,
  the statistical office of Canton Lucerne. `pv_fetch_lustat()` fetches
  one by name.

All three run through the same download cache, and every fetch prints —
and attaches to its result — where the data comes from and under which
terms you may use it. That message is not decoration: open government
data is almost always *open with obligations*, and the obligations
travel with the data frame.

## Finding a dataset

Start at the catalogue. A search returns a compact data frame — one row
per dataset, with the publisher, the licence terms, and the file formats
on offer:

```{r search, eval = FALSE}
pv_search_opendata("Finanzausgleich Luzern", rows = 20)
```

The portal's search is generous — it matches any of the words — so scan
the `title` and `id` columns for the dataset you actually mean. The `id`
is the slug the fetcher takes, and the `licence` column is worth a
glance before you commit to anything; more on the grades below.

## Fetching it

`pv_fetch_opendata()` looks the dataset up, picks its first CSV resource
(or the one you name with `resource =` — a resource id, URL, or file
name; datasets often bundle several tables), downloads it from the
publisher's server, and parses it:

```{r fetch-opendata, eval = FALSE}
fiscal <- pv_fetch_opendata("finanzausgleich-kanton-luzern",
                            resource = "fa-lu-ra.csv")
#> Source: LUSTAT Statistics Lucerne, "Equalisation of resources" via
#>   opendata.swiss (dataset "finanzausgleich-kanton-luzern")
#> Licence: opendata.swiss OPEN BY ASK terms - free use with source
#>   citation ("Quelle: LUSTAT Statistics Lucerne"); commercial use
#>   requires the data owner's permission.
```

One rule is built in: **only openly licensed resources are delivered**.
A resource published under a non-commercial or no-derivatives licence —
or with no stated licence at all — is refused with a message pointing
you at the dataset page, instead of being silently handed over.

## What the licence lines mean

opendata.swiss grades its datasets into four terms-of-use levels. All
four are open data; they differ in what they oblige you to do:

* **OPEN** — free use, no conditions.
* **OPEN BY** — free use, but you must cite the source. Most BFS data
  is published this way; the citation is
  "Quelle: Bundesamt für Statistik".
* **OPEN ASK** — free use; *commercial* use requires the data owner's
  permission.
* **OPEN BY ASK** — both: cite the source, and ask the owner before any
  commercial use. **All LUSTAT data is published under these terms** —
  cite "Quelle: LUSTAT Statistik Luzern", and if the use is commercial,
  ask LUSTAT first.

Whatever polyviz fetches, the exact terms are printed at fetch time and
stay attached to the data frame, so a script can check them later:

```{r attributes, eval = FALSE}
attr(fiscal, "pv_source")
attr(fiscal, "pv_licence")
attr(fiscal, "pv_url")
```

The practical habit: put the citation in the `source =` line of every
chart you make from fetched data. The examples below do.

## Straight from the source portals

When you already know the dataset, skip the catalogue. LUSTAT files are
fetched by their file name on data.lustat.ch:

```{r fetch-lustat, eval = FALSE}
fa <- pv_fetch_lustat("fa-lu-ra")
#> Source: LUSTAT Statistik Luzern, dataset "fa-lu-ra" (data.lustat.ch)
#> Licence: opendata.swiss "OPEN BY ASK" terms - free use with source
#>   citation ("Quelle: LUSTAT Statistik Luzern"); commercial use requires
#>   the data owner's permission.
```

BFS dataflows on stats.swiss are fetched by their SDMX id.
`pv_fetch_bfs()` requests the labelled CSV export and tidies it: every
SDMX component becomes a column named by its lower-cased id holding the
human-readable label, the underlying code sits in a matching `<id>_code`
column, and the observation itself is `obs_value`:

```{r fetch-bfs, eval = FALSE}
pop <- pv_fetch_bfs("DF_SSV_POP_1930")
#> Source: Bundesamt für Statistik, dataflow CH1.SSV:DF_SSV_POP_1930(2026.1.0),
#>   via stats.swiss
#> Licence: opendata.swiss "OPEN BY" terms - free use, source citation
#>   required ("Quelle: Bundesamt für Statistik").
```

Two quirks of the city dataflows are worth knowing: the pseudo-city
coded `"_ST"` is the all-cities total, and the population indicator
mixes census years with current-period codes whose *label* carries the
year. Rows whose indicator code starts with `"pop_ref_period"` and whose
label is a bare four-digit year are the actual population counts:

```{r bfs-filter, eval = FALSE}
cities <- pop[startsWith(pop$ssv_pop_1930_code, "pop_ref_period") &
                grepl("^\\d{4}$", pop$ssv_pop_1930) &
                pop$ssv_swiss_city_code != "_ST", ]
```

## The cache

Every download lands in a per-user cache
(`tools::R_user_dir("polyviz", "cache")`). The first fetch hits the
network; every later fetch of the same URL is instant and works offline
— which also means a script re-run does not hammer a government server.

```{r cache, eval = FALSE}
pv_cache_status()   # one row per cached download: url, size, fetched
pv_cache_clear()    # start over

# force a fresh download of one dataset, e.g. after a yearly update
fa <- pv_fetch_lustat("fa-lu-ra", refresh = TRUE)
```

## Maps to put it on

The package bundles the country's statistical geography, processed from
the BFS "Generalisierte Gemeindegrenzen" (open use, citation
"© BFS, ThemaKart"): `pv_swiss_cantons` (26 cantons),
`pv_swiss_districts` (143 districts), `pv_swiss_lakes` (the 22 largest
lakes), and `pv_lucerne_map` (the 79 Lucerne municipalities). The one
layer too big to ship — all ~2100 Swiss municipalities — is downloaded
on first use and cached:

```{r fetch-map, eval = FALSE}
muni <- pv_fetch_map("municipalities")   # needs sf on the first call only
```

`pv_choropleth()` and `pv_bubble_map()` accept all of these by name, so
you rarely touch the layer objects themselves. `map = "cantons"` joins
on the BFS canton number (1–26), `"districts"` on the district number,
`"municipalities"` on the municipality number — the ids every Swiss
statistical dataset already carries:

```{r map-layers, eval = FALSE}
pv_choropleth(canton_data, map = "cantons", id = "canton", value = "share")
pv_bubble_map(cities, lon = "lon", lat = "lat", size = "population")
```

The major lakes are drawn automatically on any map showing the whole
country (`lakes = "auto"`); force them with `lakes = TRUE` or suppress
them with `lakes = FALSE`. And a map that is not bundled at all — a
commune you surveyed, boundaries from another country — can be handed
over as an sf object; `join_id =` names the column holding each
feature's id:

```{r sf-map, eval = FALSE}
pv_choropleth(results, map = my_sf_layer, join_id = "region_code",
              id = "region", value = "turnout")
```

## From portal to paper

The whole loop, end to end: fetch the live Lucerne fiscal-equalization
data, chart the distribution, map the geography, and save both figures
for a manuscript. (This is the live version of the bundled `pv_fiscal`;
column names are LUSTAT's own — `fa_jahr` the year, `gnr` the BFS
municipality number, `ri` the resource index.)

```{r story-fetch, eval = FALSE}
fa <- pv_fetch_lustat("fa-lu-ra")
latest <- fa[fa$fa_jahr == max(fa$fa_jahr), ]
```

(One thing worth knowing about this dataset: LUSTAT publishes a year or
two ahead, so the newest years are projections, not accounts.)

First the distribution — how unequal are the municipal tax bases?

```{r story-chart, eval = FALSE}
w1 <- pv_histogram(
  latest, x = "ri", bins = 20,
  title = "Most municipalities sit below the cantonal average",
  subtitle = sprintf("Resource index, %d (average = 100)",
                     max(fa$fa_jahr)),
  source = "Quelle: LUSTAT Statistik Luzern"
)
```

Then the geography. The `gnr` column holds official BFS municipality
numbers, and the bundled `pv_lucerne_map` boundaries carry the same ids,
so the join is automatic:

```{r story-map, eval = FALSE}
w2 <- pv_choropleth(
  latest, id = "gnr", value = "ri",
  palette = "diverging", center = 100,
  title = "Tax resources follow the lake",
  subtitle = sprintf("Resource index by municipality, %d (average = 100)",
                     max(fa$fa_jahr)),
  source = "Quelle: LUSTAT Statistik Luzern; Grenzen: BFS ThemaKart"
)
```

And onto the page — `pv_save()` renders each chart for real in headless
Chrome and writes a print-quality file (see
`vignette("embedding-charts")` for sizing and formats):

```{r story-save, eval = FALSE}
pv_save(w1, "fiscal-distribution.png")   # Word, slides
pv_save(w2, "fiscal-map.pdf")            # LaTeX, true vector
```

Both figures carry their citation in the source line, which satisfies
the "BY" half of LUSTAT's OPEN BY ASK terms. The "ASK" half is yours:
these figures are fine in a paper, a thesis, or a public report — for a
commercial product, ask LUSTAT for permission first.

## Where to go next

* The charts themselves — picking forms, options, small multiples —
  are the subject of `vignette("polyviz-eda")`.
* Export formats, print sizing, and knitting to Word or PDF are in
  `vignette("embedding-charts")`.
* Every dataset bundled with the package documents its own source and
  licence on its help page — `?pv_fiscal`, `?pv_swiss_cantons`, and
  friends.
